---
output: html_document
title: Live Widget
always_allow_html: no
editor_options: 
  chunk_output_type: console
---


# parcats

`parcats` is an html widget for `easyalluvial` that creates interactive alluvial plots.

## Install

```{r cran, eval = FALSE }
install.packages('parcats')
```


## Examples

```{r}
suppressPackageStartupMessages( require(tidyverse) )
suppressPackageStartupMessages( require(easyalluvial) )
suppressPackageStartupMessages( require(parcats) )

```


### Parcats from alluvial from data in wide format

```{r eval = TRUE, fig.width = 10, fig.height = 7}
p = alluvial_wide(mtcars2, max_variables = 5)

parcats(p, marginal_histograms = TRUE, data_input = mtcars2)

```


### Parcats from model response alluvial

Machine Learning models operate in a multidimensional space and their response is hard to visualise. Model response and partial dependency plots attempt to visualise ML models in a two dimensional space. Using alluvial plots or parrallel categories diagrams we can increase the number of dimensions.

- [Visualise model response with alluvial plots](https://www.datisticsblog.com/page/visualising-model-response-with-easyalluvial/)

Here we see the response of a random forest model if we vary the three variables with the highest importance while keeping all other features at their median/mode value.


```{r eval = TRUE,  fig.width = 10, fig.height = 7}
df = select(mtcars2, -ids )
m = randomForest::randomForest( disp ~ ., df)
imp = m$importance
dspace = get_data_space(df, imp, degree = 3)
pred = predict(m, newdata = dspace)
p = alluvial_model_response(pred, dspace, imp, degree = 3)

parcats(p, marginal_histograms = TRUE, imp = TRUE, data_input = df)

```
